---
title: "Fixed Effects Homework"
author: "Jennifer Hill, Ray Lu & Zarni Htet"
output: word_document
---
## Objective 
The goal of this exercise is to simulate data that might be used to estimate a causal effects using a fixed effects regression.

## Setting
The hypothetical setting is similar to the twins study discussed briefly in class.  We are interested in estimating the effect of birthweight on future outcomes such as IQ.  However we are concerned that birthweight is correlated with other characteristics of the children that are also associated with IQ. In particular, here we suppose that the mother's IQ is the sole confounder when identifying the effect of birthweight on subsequent IQ.  

#### Question 1. God role: simulate data.
To explore what's going on here we want to simulate both the kind of data that the researchers for that study had available to them and the data that gods would have available.  

Simulate data for 1000 observations (500 pairs of twins) for the following variables:  index for the mother (momindex), child birth weight in thousands of grams (W), child IQ at age 5 (Y), mother's IQ at time of birth (X). 

The data should adhere to the following model
  $$
  X_k \sim N(100,15^2) \\
  W_{ik} \mid X \sim N(3 + .02(X_{ik}-100), .5) \\
  Y_{ik} | X_{ik}, W_{ik} \sim N(100 + .05(X_{ik}-100) + 0 W_{ik}, 5^2)
  $$
where k indexes mothers and i indexes children within mothers.  Thus X_{ik} represents the appropriate X_k value for the ith child born to the kth mother (this is the same for both children of the same mother).  
  
```{r}
set.seed(1234)
N <- 1000
K <- 500
momid <- rep(1:K,each=2)
# X_k has length K but when used in the models will be 
# appropriately expanded to length N
Xk = rnorm(K, 100, 15)
# now the child specific vectors
W = rnorm(N, 3 + (Xk[momid]-100)*.02, .5)
Y = rnorm(N, 100 + (Xk[momid]-100)*.05 + 0*W, 5)
```

### Question 2:  God role
  a) What is the effect of birthweight on IQ at age 5?
  ** No effect! **
  b) What role does X play in this scenario?
  ** It is a confounder. **

### Question 3: Researcher role
The researcher only has available to her the mother ID, the birth weight of the child, and the IQ of the child at age 5.  Given this what might a researcher to do estimate the effect of birth weight on child IQ?  Try two approaches:
 a) A regression of child IQ on birth weight.
 b) A regression of child IQ on birth weight controlling for mom fixed effects.

```{r}
summary(lm(Y ~ W))$coef[2,]
summary(lm(Y ~ W + factor(momid)))$coef[2,]
```

### Question 4: God role
Comment on the difference between these estimates and the truth.

** The first regression ignores the mom-specific confounder which results in a biased estimate of the truth.  The second regression captures this through the fixed effects and thus should not be biased on average.  The estimate for this sample is close to the truth and, appropriately, not statistically significant. **

### Question 5: Assumptions
  a) How would you formalize the ignorability assumption required by the fixed effects strategy for identifying the causal effect?
$$  
Y(w) $\indep$ W | X, \alpha
$$
Where $\alpha$ denotes the mother fixed effects.

  b) Besides ignorability, what other assumptions are built into this fixed effects strategy?
** Linear, additive model for the response surface!! **


 