---
title: "HW3_Part A"
author: "Yongchao Zhao"
date: "9/30/2018"
output: word_document
---
## Part A: Linear Parametric form  
### Question 1: Simulate the data  

(a) Start with the marginal distribution of X. Simulate as X~N(0,1) with sample size of 1000. Set the seed to be 1234.

```{r}
# DGP
set.seed(1234)
X <- rnorm(1000, mean=0, sd=1)

summary(X)

```

(b) Look at the DGP. What role does X play?  


(c) The distribution of binary Z depends on the value of X. Therefore, the next step is to simulate Z from p(Z|X) = Binomial(p), where the vector of probabilities can vary across observations. Come up with a strategy for generating the vector Z conditional on X that forces you to create be explicit about how these probabilities are conditional on X (an inverse logit function would be one strategy but there are others). Make sure that X is significantly associated with Z and that the vector of probabilities used to draw Z doesn’t vary below .05 or above .95.  

```{r}
# Probability of treatment assignment by using inverse logit function
library(gtools)
p <- inv.logit(X, min = .05, max = .95)
summary(p)

# treatment assignment
Z <- as.factor(rbinom(1000, 1, p))
table(Z)

```

(d) The last step is to simulate Y from p(Y0,Y1|Z,X). Come up with a strategy for simulating each potential outcome with appropriate conditioning on Z and X with the following stipulations. 
* (i) Make sure that E[Y(1)|X] - E[Y(0)|X] = 5.
* (ii) Make sure that X has a linear and statistically significant relationship with the outcome.
* (iii) Finally, set your error term to have a standard deviation of 1 and allow the residual standard error to be different for the same person across potential outcomes.
* (iv) Create a data frame containing X,Y,Y0,Y1 and Z and save it for use later.

```{r}
# simulate potential outcomes assuming a linear relationship (set beta0 = 10, beta1=1, and tao =5)
Y0 <- 10 + X + rnorm(1000, 0, 1)
Y1 <- 10 + X + 5 + rnorm(1000, 0, 1)

# generate the simulated dataset
data.lin <- data.frame(X, Z, Y0, Y1)
data.lin$Y <- ifelse(data.lin$Z == 1, data.lin$Y1, data.lin$Y0) # observed outcome

head(data.lin)
```

(e) Think about the difference between the DGP used in this homework and the first DGP from previous homework (completely randomized experiment). How is the difference in the study design encoded?  


(f) Calculate the SATE from 1.d.iv     

```{r}
(SATE <- mean(data.lin$Y1 - data.lin$Y0))
```

### Question 2: Playing the role of the researcher  
(a) Estimate the treatment effect using a difference in mean outcomes across treatment groups (save it for use later).  

```{r}
(avgs <- tapply(data.lin$Y, data.lin$Z, mean))
(t_mean_diff <- avgs[[2]] - avgs[[1]])

```

(b) Estimate the treatment effect using a regression of the outcome on the treatment indicator and covariate (save it for use later). 

```{r}
summary(lm(Y ~ Z + X, data=data.lin))

(t_regreesion <- coef(lm (Y ~ Z + X, data = data.lin))[[2]])

```

(c) Create a scatter plot of X versus the observed outcome with different colors for treatment and control observations (suggested: red for treated and blue for control). If you were the researcher would be comfortable using linear regression in this setting?  

```{r}
library(ggplot2)
data.lin$Z.desc <- ifelse(data.lin$Z == 1, "1-treatment", "0-control")
ggplot(data.lin, aes(X, Y, color=Z.desc)) + geom_point() + scale_color_manual(values = c("blue","red"))
       
```

###Question 3: Exploring the properties of estimators  
(a) Create a scatter plot of X versus each potential outcome with different colors for treatment and control observations (suggested: red for Y(1) and blue for Y(0)). Is linear regression a reasonable model to estimate causal effects for the observed data set? Why or why not?  

```{r}
# reshape dataset from wide to long for plotting
library(tidyr)
data.lin_new <- data.lin
data.lin_new$ID <- 1:1000
data.lin_new_long <- gather(data.lin_new, po_outcomes, po_values, Y0:Y1, factor_key = T)
# check the reshape
head(data.lin_new, 3)
subset(data.lin_new_long, ID %in% c(1:3))

# Plot
ggplot(data.lin_new_long, aes(X, po_values, color = po_outcomes)) + geom_point() + scale_color_manual(values=c("blue", "red"))

```

(b) Find the bias of each of the estimates calculated by the researcher in Question 2 relative to SATE.  

```{r}
# bias of mean difference approach
(bias_mean_diff <- t_mean_diff - SATE)

# bias of regression approach
(bias_regression <- t_regreesion - SATE)
   
```

(c) Think harder about the practical significance of the bias by dividing this estimate by the standard deviation of the observed outcome Y.  

```{r}
# practical significance of bias of mean difference approach
bias_mean_diff/sd(data.lin$Y)

# practical significance of bias of regression approach
bias_regression/sd(data.lin$Y)

```


(d) Find the bias of each of the estimators by creating randomization distributions for each. [Hint: When creating randomization distributions remember to be careful to keep the original sample the same and only varying treatment assignment and the observed outcome.]  

```{r}
data.lin.rand <- data.lin[, c("X", "Z", "Y0", "Y1", "Y")]
data.lin.rand$ID <- 1:1000

# randomly assign the treatment
set.seed(100)
sampling <- sample(data.lin.rand$ID, 500, replace = F)
data.lin.rand$Z_rand <- as.factor(ifelse(data.lin.rand$ID %in% sampling, 1, 0))

# new observed outcome
data.lin.rand$Y_rand <- ifelse(data.lin.rand$Z_rand == 1, data.lin.rand$Y1, data.lin.rand$Y0)
summary(data.lin.rand)

# new bias of mean differences 
(avgs_new <- tapply(data.lin.rand$Y_rand, data.lin.rand$Z_rand, mean))
(t_mean_diff_new <- avgs_new[[2]] - avgs_new[[1]])
(bias_mean_diff_new <- t_mean_diff_new - SATE)

# new bias of regression
summary(lm(Y_rand ~ Z_rand + X, data = data.lin.rand))
(t_regreesion_new <- coef(lm(Y_rand ~ Z_rand + X, data = data.lin.rand))[[2]])
(bias_regression_new <- t_regreesion_new - SATE)

```

